---
title: "Reproducible Research Project2"
author: "JP Van Steerteghem"
date: "9/7/2017"
output: 
  html_document: 
    keep_md: yes
---
##The impact of severe weather events in the United States from 1950 to 2011

###SYNOPIS
In this study we investigate the impact of severe weather events in the United States using the [U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database][1].   The objective of this study was to answer the following questions:

- which types of events have been most harmful with respect to population health?
- which types of events have had the greatest economic impact?


[1]:https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2/ "U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database"

### LOADING AND PROCESSING THE DATA

The following R-packages were used in this study.
```{r}
setwd("/Users/jvanstee/datasciencecoursera/RepData_PeerAssessment2")
library(dplyr)
library(ggplot2)
library(gridExtra)
```

The data was downloaded from the course website and then decompressed.  
```{r, cache=TRUE}

  if(!file.exists("./Data")){dir.create("./Data")}
  fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  download.file(fileUrl,destfile="./Data/StormData.csv.bz2")
  stormdata <- read.csv(bzfile("./Data/StormData.csv.bz2"))

#verify
str(stormdata)
```

### WHAT WEATHER EVENTS HAVE BEEN MOST HARMFUL TO THE UNITED STATES POPULATION?

```{r}
#we subset the dataset to only contain variables significant for this analysis.
#EVTYPE = event type
#FATALITIES = fatalities
#INJURIES = injuries
#PROPDMG = property damage
#PROPDMGEXP = exponent value for property damage
#CROPDMG = crop damage
#CROPDMGEXP = exponent value for crop damage

stormdataPA2 <- select(stormdata, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP,CROPDMG, CROPDMGEXP)

#determine the top10 weather events causing fatalities
casualty_ordered_fatal <-stormdataPA2 %>% group_by(EVTYPE) %>%summarise(FATALITIES = sum(FATALITIES)) %>% arrange(desc(FATALITIES)) 
fatal_top10 <- casualty_ordered_fatal[1:10, ]

#determine the top10 weather events causing injuries
casualty_ordered_injury <-stormdataPA2 %>% group_by(EVTYPE) %>%summarise(INJURIES = sum(INJURIES)) %>% arrange(desc(INJURIES))
injury_top10 <- casualty_ordered_injury[1:10, ]

#determine top10 weather events causing casualties
casualty_total = aggregate(FATALITIES + INJURIES ~ EVTYPE, stormdataPA2, sum)
names(casualty_total) = c("EVTYPE","Totalcasualties")
casualty_ordered_total = arrange(casualty_total,desc(Totalcasualties))
casualties_top10 <- casualty_ordered_total[1:10, ]

fatal_top10
injury_top10
casualties_top10

#convert EVTYPE to factor to allow proper graphing
casualties_top10$EVTYPE <- as.character(casualties_top10$EVTYPE)
casualties_top10$EVTYPE <- factor(casualties_top10$EVTYPE, levels  =unique(casualties_top10$EVTYPE))


ggplot(casualties_top10, aes(EVTYPE, Totalcasualties))+geom_histogram(stat = "identity")+theme(axis.text.x = element_text(angle = 45,hjust = 1))+ xlab("Severe Weather Type") + 
  ggtitle("Total Casualties by Severe Weather Events in the U.S. from 1950 - 2011")
```

### WHAT WEATHER EVENTS HAVE CAUSED THE GREATES ECONOMIC DAMAGE TO THE UNITED STATES?

```{r}
## Convert the property damage and crop damage data into comparable numerical forms. 
# PROPDMGEXP and CROPDMGEXP columns record contain a multiplier for each observation where we have Hundred (H), Thousand (K), Million (M) and Billion (B).

unique(stormdataPA2$PROPDMGEXP)

stormdataPA2$PROPEXP[stormdataPA2$PROPDMGEXP == "+"] <- 0
stormdataPA2$PROPEXP[stormdataPA2$PROPDMGEXP == "-"] <- 0
stormdataPA2$PROPEXP[stormdataPA2$PROPDMGEXP == "?"] <- 0

# Assigning values for the property exponent data 
stormdataPA2$PROPEXP[stormdataPA2$PROPDMGEXP == "K"] <- 1000
stormdataPA2$PROPEXP[stormdataPA2$PROPDMGEXP == "M"] <- 1e+06
stormdataPA2$PROPEXP[stormdataPA2$PROPDMGEXP == ""] <- 1
stormdataPA2$PROPEXP[stormdataPA2$PROPDMGEXP == "B"] <- 1e+09
stormdataPA2$PROPEXP[stormdataPA2$PROPDMGEXP == "m"] <- 1e+06
stormdataPA2$PROPEXP[stormdataPA2$PROPDMGEXP == "0"] <- 1
stormdataPA2$PROPEXP[stormdataPA2$PROPDMGEXP == "5"] <- 1e+05
stormdataPA2$PROPEXP[stormdataPA2$PROPDMGEXP == "6"] <- 1e+06
stormdataPA2$PROPEXP[stormdataPA2$PROPDMGEXP == "4"] <- 10000
stormdataPA2$PROPEXP[stormdataPA2$PROPDMGEXP == "2"] <- 100
stormdataPA2$PROPEXP[stormdataPA2$PROPDMGEXP == "3"] <- 1000
stormdataPA2$PROPEXP[stormdataPA2$PROPDMGEXP == "h"] <- 100
stormdataPA2$PROPEXP[stormdataPA2$PROPDMGEXP == "7"] <- 1e+07
stormdataPA2$PROPEXP[stormdataPA2$PROPDMGEXP == "H"] <- 100
stormdataPA2$PROPEXP[stormdataPA2$PROPDMGEXP == "1"] <- 10
stormdataPA2$PROPEXP[stormdataPA2$PROPDMGEXP == "8"] <- 1e+08

# Calculating the property damage value
stormdataPA2$PROPDMGVAL <- stormdataPA2$PROPDMG * stormdataPA2$PROPEXP

unique(stormdataPA2$CROPDMGEXP)

# Assigning '0' to invalid exponent data
stormdataPA2$CROPEXP[stormdataPA2$CROPDMGEXP == "?"] <- 0

# Assigning values for the crop exponent data 
stormdataPA2$CROPEXP[stormdataPA2$CROPDMGEXP == "M"] <- 1e+06
stormdataPA2$CROPEXP[stormdataPA2$CROPDMGEXP == "K"] <- 1000
stormdataPA2$CROPEXP[stormdataPA2$CROPDMGEXP == "m"] <- 1e+06
stormdataPA2$CROPEXP[stormdataPA2$CROPDMGEXP == "B"] <- 1e+09
stormdataPA2$CROPEXP[stormdataPA2$CROPDMGEXP == "0"] <- 1
stormdataPA2$CROPEXP[stormdataPA2$CROPDMGEXP == "k"] <- 1000
stormdataPA2$CROPEXP[stormdataPA2$CROPDMGEXP == "2"] <- 100
stormdataPA2$CROPEXP[stormdataPA2$CROPDMGEXP == ""] <- 1

#calculating value for crop damage
stormdataPA2$CROPDMGVAL <- stormdataPA2$CROPDMG * stormdataPA2$CROPEXP

##determine the top10 weather events causing property damage
propdmg_impact <-stormdataPA2 %>% group_by(EVTYPE) %>%summarise(PROPDMGVAL = sum(PROPDMGVAL)) %>% arrange(desc(PROPDMGVAL)) 
propdmg_impact_top10 <- propdmg_impact[1:10, ]

###determine the top10 weather events causing crop damage
cropdmg_impact <-stormdataPA2 %>% group_by(EVTYPE) %>%summarise(CROPDMGVAL = sum(CROPDMGVAL)) %>% arrange(desc(CROPDMGVAL)) 
cropdmg_impact_top10 <- cropdmg_impact[1:10, ]

propdmg_impact_top10
cropdmg_impact_top10

#convert EVTYPE to factor to allow proper graphing
propdmg_impact_top10$EVTYPE <- as.character(propdmg_impact_top10$EVTYPE)
propdmg_impact_top10$EVTYPE <- factor(propdmg_impact_top10$EVTYPE, levels = unique(propdmg_impact_top10$EVTYPE))

cropdmg_impact_top10$EVTYPE <- as.character(cropdmg_impact_top10$EVTYPE)
cropdmg_impact_top10$EVTYPE <- factor(cropdmg_impact_top10$EVTYPE, levels  =unique(cropdmg_impact_top10$EVTYPE))

#plotting graph showing the economic impact of severe weather events
propdmg_impact_top10_plot <- ggplot(propdmg_impact_top10, aes(EVTYPE, PROPDMGVAL/(10^9)))+geom_histogram(stat = "identity")+theme(axis.text.x = element_text(angle = 45,hjust = 1))+ xlab("Severe Weather Type") +ylab("Economic Impact in $B") + 
  ggtitle("Economic impact of property damage caused by Severe Weather\n Events in the U.S.\n from 1950 - 2011")
cropdmg_impact_top10_plot <- ggplot(cropdmg_impact_top10, aes(EVTYPE, CROPDMGVAL/(10^9)))+geom_histogram(stat = "identity")+theme(axis.text.x = element_text(angle = 45,hjust = 1))+ xlab("Severe Weather Type") + ylab("Economic Impact in $B") + 
  ggtitle("Economic impact of crop damage caused by Severe Weather\n Events in the U.S.\n from 1950 - 2011")
grid.arrange(propdmg_impact_top10_plot, cropdmg_impact_top10_plot, ncol = 2, right = 4, left = 4)
```

###RESULTS
This study reveals that tornadoes have had the most harmful impact on US population health.  Over the measurement period tornadoes caused 5,633 fatalities and 91,346 injuries.

The study also reveals that the significant impact of severe weather events on the US economy.  The majority of the economic impact was due to property damage.  Floods caused almost $145B in property damage, while hurricanes caused $69B and tornadoes caused $57B in property damage.  While the economic impact of crop damage was less severe, drought still caused the most damage at $14B.